---
title: "ANLY503_HW5_Xin_Lu_p1"
author: "Xin"
date: "10/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

4. For the account with the highest number of transactions, make a time series line plot for the behavior of the account over time, including all debits and credits, the different methods, and the with the different categories.

```{r}
# install.packages("reticulate")
library(reticulate)
# use_virtualenv("myenv")
# system("virtualenv -p /usr/bin/python3 /home/andres/.virtualenvs/py3-virtualenv")
# virtualenv_create("py3-virtualenv", python = "/usr/bin/python3")
# use_virtualenv("py3-virtualenv")
# py_install("pandas", envname = "py3-virtualenv")
# https://community.rstudio.com/t/problem-installing-python-libraries-with-reticulate-py-install-error-pip-not-found/26561


# use_python("/usr/local/bin/python3", required = T)
# sys <- import("sys")
# sys$version

# create a conda environment
# conda_create('r-reticulate')

# check what conda environments are available
conda_list()
# choose one conda environement to run python
use_python("/Users/xinlu/opt/anaconda3/envs/r-reticulate/bin/python", required = T)

# confirm the python environment has been changed to the one listed above
Sys.which("python")

use_condaenv("r-reticulate")

# install packages into the conda environment
# conda_install('r-reticulate', packages = 'numpy')
# check the package has been installed
py_module_available("numpy")
# conda_install('r-reticulate', packages = 'pandas')
# conda_install('r-reticulate', packages = 'matplotlib')
# conda_install('r-reticulate', packages = 'seaborn')
py_module_available("seaborn")
```



```{python import libraries}
import pandas as pd
# import random
import numpy as np
import matplotlib.pyplot as plt
import seaborn as sns
```


```{python read data}
transactions= pd.read_csv("data/transactions.csv")
transactions.info()
```

First, I need to find the account with the highest number of transactions, which is account_id = 8261, and I extract all info from this account to expedite the processing time. Then I make sure the column data is in datetime format, and set as index. Next in order to distinguish debit from credit, I change debit to negative values.
```{python data prep}
transactions['account_id'].value_counts().sort_values(ascending=False)

transactions = transactions.loc[transactions['account_id'] == 8261]

# convert date from string to datetime
transactions['date'] = pd.to_datetime(transactions['date'])

# need to set date as index for time series plot
transactions = transactions.set_index('date').sort_index()

# update debit to negative values
# [ expression for item in list if conditional ]
transactions['amount'] = [(-1)*x if j == 'debit' else x for x,j in zip(transactions['amount'], transactions['type'])]

# show all the columns
# pd.set_option('display.max_columns', None)
# transactions.head(10)
# transactions.info()
```


From the debit and credit transaction of this account over time, we can see very clearly that the credit transactions behave very consistently over time, debit transactions on the other hand, behave kindly unpredictable over time with some very dramatic changes. The transactions kind of make sense because credit transactions can be regular deposits such as salaries or investments coming into banks, and debit transactions are withdrawal so they can vary depending on what the customer is purchasing.  

The behaviors of transactions of different methods confirm the theory above that the the bank transfers are all credit transactions that can be bi-weekly salaries deposited by the employer. The cash transactions are all debit that have some very big withdrawals, which is a bit surprising because you wouldn't expect someone to withdraw over 10,000 from the bank in cash, usually transactions of this big amount are done via transfers. Yet this account has many cash transactions over 10,000.  

From graph transactions of different categories, we can see some very consistent behaviors, such as the account has a two-year loan that pays the same amount from 1993 to 1995. The account also has a house payment that pays out the same amount every month starting from 1993.  
```{python visualizations}
plt.figure(5)
fig5 = sns.lineplot(x=transactions.index, y=transactions['amount'], data=transactions, hue=transactions['type'])
fig5.set_title("Account_Id = 8261's debit & credit transactions amounts over time")
plt.show()

plt.figure(6)
sns.lineplot(x=transactions.index, y=transactions['amount'], data=transactions, hue=transactions['method'])
plt.title("Account_Id = 8261's transaction amounts of different methods over time")
plt.show()

plt.figure(7)
sns.lineplot(x=transactions.index, y=transactions['amount'], data=transactions, hue=transactions['category'])
plt.title("Account_Id = 8261's transactions amounts of different categories over time")
plt.show()
```

